---
title: "Visualising CO2 emissions"
author: "DHJ"
date: "13 12 2021"
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true   
---

*Note*: to make the maps, you need to download a shapefile from https://www.naturalearthdata.com/    

* Either use [this link](https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/50m/cultural/ne_50m_admin_0_countries.zip)   

* Or go to the [downloads page](https://www.naturalearthdata.com/downloads) and follow this path:   
    - Medium scale data, 1:50m -> Cultural -> Admin 0 – Countries

## Packages 
```{r setup, include=FALSE}

library(ggplot2)       # plotting
library(dplyr)         # group_by, summarise, %>% and all that
library(tidyr)         # for function 'pivot_wider'
library(janitor)       # for function 'clean_names'  
library(ggeasy)        # for function 'easy_rotate_labels'  
library(visreg)        # for function 'visreg' 
library(ggrepel)       # for function 'geom_text_repel'
library(RColorBrewer)  # for color scales
library(sf)            # reading, making and writing maps  

# install.packages("recmap")
# library(recmap)

knitr::opts_chunk$set(results = 'hold')   # output all the results from a chunk together

```

### Function for displaying colors  
```{r}
display_col <- function(colors){
  n <- length(colors)
  image(1:n, 1, as.matrix(1:n), col = colors)
}
```


## Data set 'Env_Dataset_UNdata' CO2 emissions    

* Denoted as data set ´dat3a´   

```{r, fig.width=9, fig.height=9}

dat3a <- read.csv("Env_Dataset_UNdata/Carbon Dioxide Emission Estimates.csv") %>%
  clean_names()

dat3a$series %>% table()  

dat3a %>%
  filter(series == "Emissions per capita (metric tons of carbon dioxide)") %>%
  ggplot(aes(year, value)) + 
  geom_point() +
  geom_line() +
  facet_wrap(vars(co2_emission_estimates)) +
  easy_rotate_labels("x", angle = -30)
  

```

## CO2 increase and population increase  

### Get data and population size per year  
```{r}


dat3a_all <- read.csv("Env_Dataset_UNdata/Carbon Dioxide Emission Estimates.csv") %>%
  clean_names() %>%
  rename(
    Country = co2_emission_estimates
  ) %>%
  # Put data on "wide" format:
  pivot_wider(names_from = "series", values_from = "value") %>%
  rename(
    co2_emissions = `Emissions (thousand metric tons of carbon dioxide)`,
    co2_percapita = `Emissions per capita (metric tons of carbon dioxide)`
  ) %>%
  mutate(
    co2_emissions = co2_emissions/1000,
    population = co2_emissions / co2_percapita 
  )

```

### Countries with data in all years    
* Selected countries represents well over 90% of global population   
```{r}

tab <- xtabs(~Country + year, dat3a_all)

# apply(tab>0, 1, sum)  
# apply(tab>0, 2, sum)  

# Pick the countries which have data in all 8 years  
tab_sel <- tab[apply(tab>0, 1, sum) == 8,]
selected_countries <- rownames(tab_sel)

cat("Number of countries selected: \n")
length(selected_countries)

cat("\n")
cat("Total population size (in millions) of countries selected: \n")
dat3a_all %>%
  filter(Country %in% selected_countries & year == 2017) %>%
  summarise(Total_pop = sum(population)) %>%
  pull(Total_pop)

cat("Total population size (in millions) of all countries: \n")
dat3a_all %>%
  filter(year == 2017) %>%
  summarise(Total_pop = sum(population)) %>%
  pull(Total_pop)

```

### Total for all countries   
Almost all increase in CO2 is *seemingly* explained by population increase   
```{r, fig.width=8, fig.height=4}

# Sum all numbers ('where(is.double)') for every year
dat_sum <- dat3a_all %>%
  filter(Country %in% selected_countries) %>%
  group_by(year) %>%
  dplyr::summarize(across(where(is.double), .fns = sum))

# Statistical model  
mod1 <- lm(co2_emissions ~ co2_percapita + population, dat_sum) 
# summary(mod1)

# Plot result of statistical model
par(mfrow = c(1,2), mar=c(4,4,2,1))
visreg(mod1, line=list(col="red2"),
                    fill=list(col="pink"),
                    points=list(cex=1.5, pch=19),
       ylim = c(5000,30000))

# Percent explained by population alone  
mod0 <- lm(co2_emissions ~ 1, dat_sum)
mod0a <- lm(co2_emissions ~ co2_percapita, dat_sum)
mod0b <- lm(co2_emissions ~ population, dat_sum)
res0 <- sum(mod0$residuals^2)
res0a <- sum(mod0a$residuals^2)
res0b <- sum(mod0b$residuals^2)
res1 <- sum(mod1$residuals^2)
cat("Percent explained by co2 per capita alone: \n")  
(res0 - res0a)/res0*100
cat("Percent explained by population alone: \n")  
(res0 - res0b)/res0*100
cat("Percent explained by population + co2 per capita: \n")  
(res0 - res1)/res0*100

cat("\nANOVA:\n")
anova(mod1)

```


### For China alone  

* Decided that a statistical model is overkill (we *know* that co2_emissions = population*co2_percapita)
* Calculate all values as percentage of the 1975 value (in other words, set 1975 = 100)  
* For China, the increase in CO2 is mostly due to the increase in *CO2 per capita*, *not* population increase    
* Note: coulnd to this on log scale, as then the 

```{r, fig.width=8, fig.height=4}

# Function for value as percentage of the first ()   
value_rel <- function(x) {x/head(x,1)*100}

dat_sel <- dat3a_all %>%
  filter(Country %in% "China") %>%
  dplyr::summarize(across(where(is.double), .fns = value_rel))
dat_sel

percap <- tail(dat_sel$co2_percapita,1)/tail(dat_sel$co2_emissions,1)
popsize <- tail(dat_sel$population,1)/tail(dat_sel$co2_emissions,1)

cat("'Fraction' of CO2 increased explained by per capita CO2 increase: \n ")
percap/(percap + popsize)

```

#### Trajectory for China  
```{r}

ggplot(dat_sel, aes(population, co2_percapita)) +
  geom_line() +
  geom_point(aes(size = co2_emissions, color = co2_emissions)) + 
  scale_color_gradient("Total CO2 emissions") +
  scale_size("Total CO2 emissions") +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "Population size (1970 = 100)",
    y = "CO2 per capita (1970 = 100)",
    title = "China, emission and population trajectory 1975-2017"
  )

```


### For Sweden    

* CO decrease, while the population has increased  

```{r, fig.width=8, fig.height=4}

change_rel <- function(x) {(x - head(x,1))/head(x,1)*100}
value_rel <- function(x) {x/head(x,1)*100}

dat_sel <- dat3a_all %>%
  filter(Country %in% "Sweden") %>%
  dplyr::summarize(across(where(is.double), .fns = value_rel))
dat_sel

```

### For all countries    

* Works for countries where  both population size and co2 emissions has increased  
* For countries where co2 emissions has decreased (e.g. Sweden) it's not meaningful    
* This approach is abandoned  

```{r, fig.width=8, fig.height=4}

value_rel <- function(x) {x/head(x,1)*100}

spliteffects_one_country <- function(country){

  dat_sel <- dat3a_all %>%
    filter(Country %in% country) %>%
    dplyr::summarize(across(where(is.double), .fns = value_rel))
  
  result <- data.frame(
    Country = country,
    co2_emissions_rel = tail(dat_sel$co2_emissions,1),
    co2_percapita_rel = tail(dat_sel$co2_percapita,1),
    population_rel = tail(dat_sel$population,1)
  )
  
  result$percap_rat <- with(result, co2_percapita_rel/co2_emissions_rel)
  result$popsize_rat <- with(result, population_rel/co2_emissions_rel) 
  
  result$percap_perc <- with(result, percap_rat/(percap_rat + popsize_rat)*100)
  result$popsize_perc <- with(result, popsize_rat/(percap_rat + popsize_rat)*100)
  
  result
  
}

# test
spliteffects_one_country("China")
spliteffects_one_country("Sweden")

# For all countries   
spliteffects <- selected_countries %>% purrr::map_dfr(spliteffects_one_country)

```

## CO2 increase and population increase - maps   

### Get map   

* Shapefile from https://www.naturalearthdata.com/  

```{r}

map_countries <- st_read("ne_50m_admin_0_countries/ne_50m_admin_0_countries.shp")

```

### Make 'dat3a_2017'   
* Contains   
    - emissions (total and per capita) for 2017  
    - scaled emissions (total and per capita) + population for 2017 (all scaled to 1975 = 100)      
    - difference between total scaled emissions and total scaled population     
    - difference between total scaled emissions and total scaled per capita emissions       
```{r}

# dat3a %>%
#   filter(series == "Emissions (thousand metric tons of carbon dioxide)") %>% xtabs(~year, .) 

# 2017 only  
dat3a_2017 <- read.csv("Env_Dataset_UNdata/Carbon Dioxide Emission Estimates.csv") %>%
  clean_names() %>%
  filter(year == 2017) %>%
  rename(
    Country = co2_emission_estimates
  ) %>%
  pivot_wider(names_from = "series", values_from = "value") %>%
  rename(
    co2_emissions_2017 = `Emissions (thousand metric tons of carbon dioxide)`,
    co2_percapita_2017 = `Emissions per capita (metric tons of carbon dioxide)`
  ) %>%
  mutate(
    population_2017 = co2_emissions_2017 / co2_percapita_2017 /1000
  )   


# All years, on wide format  
dat3a_wide <- read.csv("Env_Dataset_UNdata/Carbon Dioxide Emission Estimates.csv") %>%
  clean_names() %>%
  rename(Country = co2_emission_estimates) %>%
  pivot_wider(names_from = "series", values_from = "value") %>%
  rename(
    co2_emissions = `Emissions (thousand metric tons of carbon dioxide)`,
    co2_percapita = `Emissions per capita (metric tons of carbon dioxide)`
  ) %>%
  mutate(
    co2_emissions = co2_emissions/1000,
    population = co2_emissions / co2_percapita
  ) 
  

# Function for getting 2017 value (tail) relative to
#   1975 value (head)
value_rel_2017 = function(x) {tail(x,1)/head(x,1)*100}

# Get relative values  
dat3a_addition <- dat3a_wide %>%  
  filter(Country %in% selected_countries) %>%
  arrange(Country, year) %>%
  group_by(Country) %>%
  dplyr::summarize(across(where(is.double), 
                          .fns = value_rel_2017,
                          .names = "{.col}_scaled"), 
                   .groups = "drop") %>%
  # Add difference  
  mutate(
    co2_dev_population = co2_emissions_scaled - population_scaled,
    co2_dev_percapita = co2_emissions_scaled - co2_percapita_scaled
    )

# 
dat3a_2017 <- dat3a_2017 %>%
  mutate(
    NAME = case_when(
      grepl("Brunei", Country) ~ "Brunei",
      grepl("Bolivia", Country) ~ "Bolivia",
      grepl("Iran", Country) ~ "Iran",
      grepl("Tanzania", Country) ~ "Tanzania",
      grepl("Venezuela", Country) ~ "Venezuela",
      grepl("Viet Nam", Country) ~ "Vietnam",
      grepl("Moldova", Country) ~ "Moldova",
      grepl("Syria", Country) ~ "Syria",
      grepl("Ivoire", Country) ~ "Côte d'Ivoire",
      grepl("Bosnia", Country) ~ "Bosnia and Herz.",
      grepl("Syria", Country) ~ "Syria",
      Country == "Russian Federation" ~ "Russia",
      Country == "Republic of Korea" ~ "South Korea",
      Country == "Dem. People's Rep. Korea" ~ "North Korea",
      Country == "Dem. Rep. of the Congo" ~ "Dem. Rep. Congo",
      Country == "South Sudan" ~ "S. Sudan",
      TRUE ~ Country)
  ) %>%
  left_join(dat3a_addition, by = c("NAME" = "Country"))

dat3a_2017 %>%
  arrange(desc(co2_emissions_2017)) %>% 
  head(10) %>%
  select(Country, co2_emissions_2017, co2_emissions_scaled)   


# keep only those with data
dat3a_2017 <- dat3a_2017 %>%
  filter(!is.na(co2_emissions_scaled))  

```

```{r}

dat3a_2017$co2_emissions_scaled %>% quantile()  
dat3a_2017$co2_emissions_scaled %>% quantile(seq(0,1,0.15))  

display.brewer.pal(7, name = "RdYlGn")
cols <- rev(RColorBrewer::brewer.pal(7, name = "RdYlGn")[c(1:3,6)])
cols <- c(cols, "grey40")
display_col(cols)

```



### Plot scaled emissions + population for 2017   
```{r, fig.width=8, fig.height=5}

# Find intervals for color scale 
# dat3a_2017$co2_emissions_scaled %>% quantile()  
# dat3a_2017$co2_emissions_scaled %>% quantile(seq(0,1,0.15))  

# display.brewer.all()
# cols <- rev(RColorBrewer::brewer.pal(11, name = "RdYlGn")[1:7])
#display_col(cols)

dat_countries_with_labels <- bind_rows(
  # The 10 biggest emittors in 2017
  dat3a_2017 %>% arrange(co2_emissions_2017) %>% tail(10),
  # The top 3 with biggest increase in per capita CO2  
  dat3a_2017 %>% arrange(co2_percapita_scaled) %>% tail(3),
  # The top 3 with biggest decrease in per capita CO2
  dat3a_2017 %>% arrange(co2_percapita_scaled) %>% head(3),
  # The top 1 with biggest increase in population size  
  dat3a_2017 %>% arrange(population_scaled) %>% tail(1),
  # The top 1 with biggest decrease in population size  
  dat3a_2017 %>% arrange(population_scaled) %>% head(1),
  # Norway
  dat3a_2017 %>% filter(Country == "Norway")
)

dat3a_2017 %>%
  # Change co2_emissions_scaled for continuous to categorized, for color scale:
  mutate(co2_emissions_scaled = cut(co2_emissions_scaled, breaks = c(47,100,250,500,800,9069))) %>%
  # Plotting:
  ggplot(aes(population_scaled, co2_percapita_scaled, group = Country)) +
  geom_point(aes(size = co2_emissions_2017, fill = co2_emissions_scaled), pch = 21) + 
  scale_size("Total CO2 emissions\n(kilotonnes)") +
  scale_fill_manual("Total CO2 emissions\n(1975 = 100)", values = cols) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_text_repel(
    data = dat_countries_with_labels,
    aes(label = Country), color = "red3",
    box.padding = .6, point.padding = 0.5, min.segment.length = 0, nudge_x = 0.15, nudge_y = 0.05
  ) +
  geom_hline(yintercept = 100, linetype = "dashed", size = 1, color = "grey40") +
  annotate("text", x = rep(1000,2), y = rep(100,2), 
           label = c("Increase since 1975", "Decrease since 1975"),
           vjust = c(-0.5, 1.2)) +
  labs(
    x = "Population size (1975 = 100)",
    y = "CO2 per capita (1975 = 100)",
    title = "2017 emissions + population, scaled by 1975 values (1975 = 100)",
    subtitle = "Note: log scale on both axes. Text shows the 10 largest emittors + some other extremes"
  )
# hjust = 0, nudge_x = 0.01, 

```

### Plot unscaled emissions + population for 2017   
```{r, fig.width=8, fig.height=5}

# Find intervals for color scale 
# dat3a_2017$co2_emissions_scaled %>% quantile()  
# dat3a_2017$co2_emissions_scaled %>% quantile(seq(0,1,0.15))  

# display.brewer.all()
# cols <- rev(RColorBrewer::brewer.pal(11, name = "RdYlGn")[1:7])
#display_col(cols)

dat_countries_with_labels <- bind_rows(
  # The 10 biggest emittors in 2017
  dat3a_2017 %>% arrange(co2_emissions_2017) %>% tail(10),
  # The top 3 with biggest increase in per capita CO2  
  dat3a_2017 %>% arrange(co2_percapita_scaled) %>% tail(3),
  # The top 3 with biggest decrease in per capita CO2
  dat3a_2017 %>% arrange(co2_percapita_scaled) %>% head(3),
  # The top 1 with biggest increase in population size  
  dat3a_2017 %>% arrange(population_scaled) %>% tail(1),
  # The top 1 with biggest decrease in population size  
  dat3a_2017 %>% arrange(population_scaled) %>% head(1),
  # Norway
  dat3a_2017 %>% filter(Country == "Norway")
)

dat3a_2017 %>%
  mutate(co2_emissions_scaled = cut(co2_emissions_scaled, breaks = c(47,100,250,500,800,9069))) %>%
  ggplot(aes(population_2017, co2_percapita_2017, group = Country)) +
  geom_point(aes(size = co2_emissions_2017, fill = co2_emissions_scaled), pch = 21) + 
  scale_size("Total CO2 emissions\n(kilotonnes)") +
  scale_fill_manual("Total CO2 emissions\n(1975 = 100)", values = cols) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_text_repel(
    data = dat_countries_with_labels,
    aes(label = Country), color = "red3",
    box.padding = .6, point.padding = 0.5, min.segment.length = 0, nudge_x = 0.15, nudge_y = 0.05
  ) +
  labs(
    x = "Population size (millions)",
    y = "CO2 per capita (tons)",
    title = "2017 emissions + population, absolute values",
    subtitle = "Note: log scale on both axes. Text shows the 10 largest emittors + some other extremes"
  )
# hjust = 0, nudge_x = 0.01, 

```

## Make map data  
```{r}

map_emissions <- map_countries %>%
  select(NAME) %>%
  left_join(dat3a_2017, by = "NAME") %>%
  select(-Country)


# Check country names  
n1 <- unique(dat3a_2017$NAME)
n2 <- map_emissions %>% filter(!is.na(co2_emissions_2017)) %>% pull(NAME)
n3 <- map_emissions %>% pull(NAME)

# Check country names in CO2 data not found in map data 
n1[!n1 %in% n2]  

```


### Save as Shape and GeoJSON  
```{r}

foldername <- "map_emissions"
dir.create(foldername)

# Shape
sf::write_sf(map_emissions, paste0(foldername, "/emissions_changes.shp"))

# GeoJSON
sf::write_sf(map_emissions, paste0(foldername, "/emissions_changes.GeoJSON"))

```

## Plot CO2 emission changes      

### Increase relative to population increase
```{r}

mapdata <- read_sf("nc_emissions_2017F/emissions_changes.GeoJSON")

var <- "co2_dev_population"
quantile(mapdata[[var]], na.rm = TRUE)  

brks <- c(-250, -200, -150, -100, -50, 0, 100, 200, 500, 2000, 10000)
mapdata$col1 <- cut(mapdata[[var]], 
                   breaks = brks)

# RColorBrewer::brewer.pal(11, name = "RdYlBu")

plot(mapdata["col1"], pal = rev(RColorBrewer::brewer.pal(11, name = "RdYlBu")),
     main = "How much has CO2 increased relative to population increase?")


```

#### Plot as ggplot

```{r}

# perform transformation on modified version of world dataset
mapdata2 <- st_transform(mapdata, 
                               crs = '+proj=robin +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs')

gg <- ggplot(mapdata2) +
  geom_sf(aes(fill = col1)) +
  scale_fill_brewer("CO2 increase -\npopulation increase", palette = "RdYlBu", direction = -1) +
  labs(title = "How much has CO2 increased relative to population increase?",
       subtitle = "Negative numbers = CO2 increase < population increase") +
  theme_dark()
  
ggsave("co_rel_popul.png", gg, width = 12, height = 9, dpi = 400)

gg

```



### Increase relative to per capita CO2 increase   
```{r}


mapdata <- read_sf("nc_emissions_2017F/emissions_changes.GeoJSON")

var <- "co2_dev_percapita"
#quantile(mapdata[[var]], na.rm = TRUE)  

brks <- c(-250, -200, -150, -100, -50, 0, 100, 200, 500, 2000, 10000)
mapdata$col2 <- cut(mapdata[[var]], 
                   breaks = brks)

# RColorBrewer::brewer.pal(11, name = "RdYlBu")

plot(mapdata["col2"], pal = rev(RColorBrewer::brewer.pal(11, name = "RdYlBu")),
     main = "How much has CO2 increased relative to co2 per capita increase?")


```

#### Plot as ggplot

```{r}

# perform transformation on modified version of world dataset
mapdata2 <- st_transform(mapdata, 
                               crs = '+proj=robin +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs')

gg <- ggplot(mapdata2) +
  geom_sf(aes(fill = col2)) +
  scale_fill_brewer("CO2 increase -\npopulation increase", palette = "RdYlBu", direction = -1) +
  labs(title = "How much has CO2 increased relative to co2 per capita increase?",
       subtitle = "Negative numbers = CO2 increase < population increase") +
  theme_dark()

ggsave("co_rel_procap.png", gg, width = 12, height = 9, dpi = 400)

gg


```


